Modelling the emergence of whisker barrels

### 2020-01-27 paper/briefpaper/brief.tex
A few more tweaks

Software for modelling the cortical sheet as a hexagonal lattice with no-flux
boundary conditions and an arbitrary boundary shape, for computing the flux
using a finite difference method, for numerical integration by the
fourth-order Runge-Kutta method, and for reproducing these results is
available at \url{https://github.com/ABRG-Models/BarrelEmerge}.

@@

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density \cmnt{of unconnected
  branches} at $t=0$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. \cmnt{Note that this
  operation is local to individual afferent projections.}

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.~\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time (Eq.~\ref{eq:norm}).

@@

The cortex was modelled as a two dimensional hexagonal lattice which
simplifies the computation of the 2D Laplacian. Within a boundary traced from
rat primary barrel cortex (Fig.\,1A) we set the hex-to-hex distance
$d=0.03$\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) \equiv \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right).
\end{equation}
%
We can define the time-independent sum of the guidance gradients
$\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla \rho_j(\mb{x})$
and then apply vector calculus identities to compute
$\nabla\vcdot\mb{J}_i$. Because the divergence operator is distributive,
Eq.~\ref{eq:divJ} can be expanded (dropping references to $\mb{x}$ and $t$ for
clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) - \nabla\vcdot\big(a_i \mb{g}_i\big).
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
,
\end{equation}
%
which has three elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i(\mb{x},t)$); ii)~a time-independent modulator of $a_i(\mb{x},t)$ (because
$\nabla\vcdot\mb{g}_i(\mb{x})$ is a time-independent static field); and iii)~the
scalar product of the static vector field $\mb{g}_i$ and the gradient of
$a_i$. Each of the divergences can be simplified by means of Gauss's Theorem
following \cite{lee_hexagonal_2014}.

(i) The computation of the mean value of the Laplacian across the area of one
hexagon of area $\Omega$ located at position $\mb{p}_0$, with neighbours at
positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \iint_A \dvrg a_i(\mb{x}, t) dxdy & = \frac{1}{\Omega} \oint_c \frac{\partial a_i}{\partial \hat{\mb{n}}} d\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{mid} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big)
\end{split}
\end{equation}
%
where $d$ is the centre-to-centre distance between hexes in the grid, $v =
d/\sqrt{3}$ is the length of each of the 6 edges of the hexagon's perimeter
and $d\gamma$ is an infinitessimally small distance along the perimeter of the
hexagon. The area, $\Omega$, of each hexagon in the lattice is
$\frac{\sqrt{3}}{2}d^2$.

@@

\begin{equation} \label{eq:divg}
\begin{split}
%
\frac{1}{\Omega} \iint_A a_i\nabla\vcdot\mb{g}\;dxdy & = \frac{a_i(\mb{p}_0)}{\Omega}  \oint_c \mb{g}\vcdot d\hat{\mathbf{n}} \\
%
& \approx \frac{a_i(\mb{p}_0,t)}{\Omega} \sum_{j=1}^{6} \frac{\mb{g}_j + \mb{g}_0}{2}\vcdot \hat{\mb{n}}\;v \\
%
& = \frac{2 a_i(\mb{p}_0,t)v}{\sqrt{3}d^2} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
%%& = \frac{2 a_i(\mb{p}_0,t)d}{\sqrt{3}d^2\sqrt{3}} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
\Rightarrow \langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle & \approx \frac{a_i(\mb{p}_0,t)}{3d} \bigg( \sum_{j=1}^{6} \big(\mb{g}_j^x + \mb{g}_0^x\big) \cos \big(\frac{\pi}{3}(j-1)\big) + \sum_{j=1}^{6} \big({\mb{g}_j^y + \mb{g}_0^y}\big) \sin \big(\frac{\pi}{3}(j-1)\big) \bigg). \\
%
\end{split}
\end{equation}
%
Both this last expression, and the final expression of Eq.\,\ref{eq:lapl} are
easily computed, comprising short sums of nearest-neighbour relationships.

iii) The final term in Eq.~\ref{eq:divJExpanded} is the scalar product of two
vector fields which is straightforward to compute from their Cartesian
components.

The advantage of separating the computation of Eq.~\ref{eq:divJ} in (i), (ii)
\& (iii) is that we can ensure that the no-flux boundary condition is
fulfilled. On the boundary, the contribution to $\mb{J}$ resulting from the
first term of Eq.~\ref{eq:divJExpanded} can be held at 0 by the ghost cell
method (in which a hex outside the boundary containing the same value as the
hex inside the boundary is imagined such that the flux of $J$ across the
boundary is 0). Then, $\mb{g}_i(\mb{x})$ can be tailored so that it, and its
normal derivative approach 0 at the boundary, ensuring that the second and
third terms of Eq.~\ref{eq:divJExpanded} also contribute nothing to
$\mb{J}$. This is achieved by applying to $\mb{g}(\mb{x})$ a sharp logistic
function of the distance from $\mb{x}$ to the boundary.

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge}. The computations described
in (i), (ii) and (iii) may be found in the code \code{compute\_divJ()} as
\code{term1}, \code{term2} and \code{term3}, respectively.

>>>>

@@

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density \cmnt{of unconnected
  branches} at $t=0$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. \cmnt{Note that this
  operation is local to individual afferent projections.}

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time (Eq.\,\ref{eq:norm}).

@@

The cortex was modelled as a two dimensional hexagonal lattice which
simplifies the computation of the 2D Laplacian. Within a boundary traced from
rat primary barrel cortex (Fig.\,1A) we set the hex-to-hex distance
$d$ to 0.03\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right).
\end{equation}
%
We can note that the sum of the guidance gradients is time-independent and
define $\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla
\rho_j(\mb{x})$ and then apply vector calculus identities to compute
$\nabla\vcdot\mb{J}_i$. Because the divergence operator is distributive,
Eq.\,\ref{eq:divJ} can be expanded (dropping references to $\mb{x}$ and $t$
for clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) - \nabla\vcdot\big(a_i \mb{g}_i\big).
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
,
\end{equation}
%
which has three elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i$); ii)~a time-independent modulator of $a_i$ (because
$\nabla\vcdot\mb{g}_i$ is a time-independent static field); and iii)~the
scalar product of the static vector field $\mb{g}_i$ and the gradient of
$a_i$. Each of the divergences can be simplified by means of Gauss's Theorem
following \cite{lee_hexagonal_2014}.

(i) The computation of the mean value of the Laplacian across one hexagon of
area $\Omega = \frac{\sqrt{3}}{2}d^2$, located at position $\mb{p}_0$, with
neighbours at positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \iint_\Omega \dvrg a_i(\mb{x}, t) dxdy & = \frac{1}{\Omega} \oint_c \frac{\partial a_i}{\partial \hat{\mb{n}}} d\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{mid} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big)
\end{split}
\end{equation}
%
where $d$ is the centre-to-centre distance between hexes in the grid, $v =
d/\sqrt{3}$ is the length of each of the 6 edges of the hexagon's perimeter
and $d\gamma$ is an infinitessimally small distance along the perimeter of the
hexagon.

@@

\begin{equation} \label{eq:divg}
\begin{split}
%
\frac{1}{\Omega} \iint_\Omega a_i\nabla\vcdot\mb{g}\;dxdy & = \frac{a_i(\mb{p}_0,t)}{\Omega}  \oint_c \mb{g}\vcdot d\hat{\mathbf{n}} \\
%
& \approx \frac{a_i(\mb{p}_0,t)}{\Omega} \sum_{j=1}^{6} \frac{\mb{g}_j + \mb{g}_0}{2}\vcdot \hat{\mb{n}}\;v \\
%
& = \frac{2 a_i(\mb{p}_0,t)v}{\sqrt{3}d^2} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
%%& = \frac{2 a_i(\mb{p}_0,t)d}{\sqrt{3}d^2\sqrt{3}} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
\Rightarrow \langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle & \approx \frac{a_i(\mb{p}_0,t)}{3d} \bigg( \sum_{j=1}^{6} \big(\mb{g}_j^x + \mb{g}_0^x\big) \cos \big(\frac{\pi}{3}(j-1)\big) + \sum_{j=1}^{6} \big({\mb{g}_j^y + \mb{g}_0^y}\big) \sin \big(\frac{\pi}{3}(j-1)\big) \bigg). \\
%
\end{split}
\end{equation}
%
Both this last expression, and the final expression of Eq.\,\ref{eq:lapl} are
easily computed, comprising short sums of nearest-neighbour relationships.

iii) The final term in Eq.\,\ref{eq:divJExpanded} is the scalar product of two
vector fields which is straightforward to compute from their Cartesian
components.

By separating the computation of Eq.\,\ref{eq:divJ} into parts (i), (ii) \&
(iii), we can ensure that the no-flux boundary condition,
$\mb{J}_i(\mb{x},t)\big\rvert_{boundary} = 0$, is fulfilled. On the boundary,
the contribution to $\mb{J}$ resulting from the first term of
Eq.\,\ref{eq:divJExpanded} can be fixed to 0 by the `ghost cell method' in
which, during the evaluation of (i), a hex outside the boundary containing the
same value as the hex inside the boundary is imagined to exist such that the
flux of $\mb{J}$ across the boundary is 0. Then, $\mb{g}_i(\mb{x})$ can be
tailored so that it, and its normal derivative approach 0 at the boundary,
ensuring that the second and third terms of Eq.\,\ref{eq:divJExpanded} also
contribute nothing to $\mb{J}$. This is achieved by applying to
$\mb{g}_i(\mb{x})$ a sharp logistic function of the distance from $\mb{x}$ to
the boundary.

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife_submission1}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James::compute\_divJ()} as \code{term1}, \code{term2} and \code{term3},
respectively.
